% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% Theory %%
% Conclusion
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
\label{sec:theory:conclu}

\lettrine{T}{his} chapter presented a theoretical model to analyze the behavior of \glspl{redC} when processing sparse matrices, specifically focusing on the \acrshort{op} \acrshort{spmv} kernel.
A C program implementing a hybrid probabilistic/fast simulation model was developed, and we demonstrated how the structure of \gls{banded} matrices can be leveraged to optimize cache performance.

Our analysis highlighted the importance of separating the in-\gls{band} and out-of-\gls{band} regions of the matrix into dedicated cache regions.
For the in-\gls{band} region, a \gls{set}-associative, line-wise cache configuration ensures efficient data reuse by maintaining a rotating \gls{eviction} pattern, minimizing unnecessary memory transactions.
For the out-of-\gls{band} region, a fully-associative, element-wise cache with an element-wise memory interface significantly reduces memory traffic by avoiding the transfer of mostly empty cache-lines.

The theoretical insights gained from this model validate the architectural choices of \acrlong{spdc}, particularly the use of distinct regions for denser and sparser data.
These findings provide a robust foundation for designing and sizing cache regions to minimize memory traffic, ultimately improving the efficiency of sparse matrix computations.
In the next chapter, we will delve into the microarchitectural details of \acrlong{spdc}, translating these theoretical insights into practical hardware design.

\bigskip\bigskip

\begin{highlight}{Takeaways}{Mulberry}
    \colorbf{Mulberry}{Simulating \emph{reduction caches} and SpDCache} \\
    To study the concepts presented in this thesis, three models were developed with varying levels of abstraction. Starting from the most abstract model:
    \begin{itemize}
        \item \textbf{\Gls{highLevel} probabilistic/fast model in C} (\ref{sec:theory:cmodel}):
        \begin{itemize}
            \item fast simulation ($\sim$10\textsuperscript{5} $nnz/s$), using generic matrix density distributions, easy parameter exploration, deep understanding of cache behaviors, but,
            \item very \gls{highLevel} (some approximations and hypothesis), limited on \gls{eviction} policies.
        \end{itemize}
        \item \textbf{Transaction level model in Python} (\ref{sec:spdc:eval}):
        \begin{itemize}
            \item faster than a full RTL simulation ($\sim$10\textsuperscript{3} $nnz/s$), accurate performance on memory traffic, functional testing, but,
            \item slower than the probabilistic model, demands real matrices as input, two factors that alleviate the parameter exploration capabilities, no latency performance compared to RTL simulation.
        \end{itemize}
        \item \textbf{RTL implementation} (next chapter):
        \begin{itemize}
            \item accurate results in both memory traffic and latencies, hardware testing, but,
            \item slow simulation ($\sim$10 $nnz/s$), demands real matrices in input, time-consuming implementation.
        \end{itemize}
    \end{itemize}

    \bigskip

    \colorbf{Mulberry}{Validation and sizing of SpDCache three regions}
    \begin{itemize}
        \item \textbf{\acrshort{grc}}: can take advantage of prefetching methods, size depends on $\mathit{nnz}_c$.
        \item \textbf{\acrshort{drc}}: must respect the relationship $s_C \geqslant w_B + l - 1$ (Equation~\ref{eq:theory:in_band_cond}).
        \item \textbf{\acrshort{srt}}: internal storage and memory interface benefit from element-wise granularity, benefits from a larger size but must achieve a compromise with hardware cost.
    \end{itemize}
    To simplify address decoding, the initial region sizing for \acrlong{spdc} was chosen somewhat arbitrarily to be \textbf{25/50/25\%} for the \acrshort{grc}, \acrshort{drc} and \acrshort{srt}, respectively.
    Our analysis later confirmed that this partitioning is actually close to optimal for memory traffic performance.
\end{highlight}

\clearpage