% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% Theory %%
% GRC
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Isolating the Input Reads in a Generic Cache}
\label{sec:theory:isolate_input_reads}

\lettrine{W}{hen} computing an \acrshort{op} \acrshort{spmv} kernel, all the \glspl{irrAccess} come from updating the output vector $c$.
Indeed, the input matrix $A$ and vector $b$ are read sequentially, and can thus be efficiently cached by a generic cache that naturally exploits spatial locality.

With the matrix $A$ stored in \acrshort{csc} format, each iteration of the \acrshort{op} algorithm necessitates reading four tables: $\mathit{A.ptr}$, $\mathit{A.idx}$, $\mathit{A.val}$ and $b$.
In the same iteration, a fifth table, $c$, is updated ``randomly''.
If we consider cache-lines of size $l$ elements, then each cache-line read of the $\mathit{A.ptr}$ table brings $l$ consecutive entries, which are pointers to $l$ consecutive columns of the matrix.
Similarly, each cache-line read of the $b$ table brings $l$ consecutive entries, which will be multiplied by the non-zero elements of $l$ consecutive columns of the matrix.
These reads are needed in the outer-loop of the \acrshort{op} algorithm.
In the inner-loop, each cache-line read of the $\mathit{A.idx}$ and $\mathit{A.val}$ tables brings $l$ consecutive non-zero elements of the matrix to process, which will generate $l$ ``random'' updates to the output vector $c$, thus $l$ different cache-lines of $c$ could be accessed in the worst case.

To avoid unnecessary \glspl{eviction}, at each iteration, the generic cache should, obviously, keep $\mathit{A.ptr}$, $\mathit{A.idx}$, $\mathit{A.val}$ and $b$ cache-lines up until they are no longer needed.
That means $l$ iterations for $\mathit{A.idx}$ and $\mathit{A.val}$, and around $l \times \mathit{nnz}_c$ iterations for $\mathit{A.ptr}$ and $b$.
However, the cache must also keep the cache-lines of $c$ that are being updated, which, over the course of $l \times \mathit{nnz}_c$ iterations, can necessitate up to the same number of different cache-lines in the worst case, on unpredictable addresses.
Thus, it would be difficult for a generic cache to keep all the useful data until it is no longer needed.
$\mathit{A.ptr}$ and $b$ cache-lines, which are accessed once every $\mathit{nnz}_c$ iterations, could easily be evicted by a \acrshort{lru} policy which is unaware of their future use.
By isolating the input reads in a dedicated cache, as proposed with \acrlong{spdc} and the \gls{redC} baseline in the previous chapter, we greatly increase the chances that all the cache-lines of $A$ and $b$ remain in the cache until they are no longer needed.
With a dedicated cache region for the inputs, the \acrshort{grc} introduced in previous chapter, only the four sequentially read tables need to be considered.
These tables are read only once, and isolating them in the \acrshort{grc} decreases the risk of undesired \glspl{eviction}.

With a \acrshort{lru} policy, cache-lines holding $\mathit{A.ptr}$ and $b$ are more at risk of being evicted than $\mathit{A.idx}$ and $\mathit{A.val}$ because they are accessed less frequently, only once every $\sim \frac{\mathit{nnz}_c}{l}$ times per column.
$\mathit{A.idx}$ and $\mathit{A.val}$ are accessed every iteration of the inner-loop.
To avoid premature \glspl{eviction} of $\mathit{A.ptr}$ and $b$, the \acrshort{grc} must be sized to handle the $\frac{\mathit{nnz}_c}{l}$ cache-line reads which occur while computing a column.
In Equation~\ref{eq:theory:grc_size}, we show the minimum size of the \acrshort{grc} in elements needed to avoid unnecessary \glspl{eviction} during the traversal of the columns, thus optimizing memory traffic.

\begin{equation}
    s_{C_\mathit{min}}(\mathit{GRC}) = 2 \times l + 2 \times \mathit{nnz}_c
    \label{eq:theory:grc_size}
\end{equation}

This relationship is not exact, as it does not consider address alignment effects nor temporal variability in the downstream memory, but it gives a good approximation of the minimum size needed for the \acrshort{grc} to avoid unnecessary \glspl{eviction}.
It explains why, in the previous chapter, the total memory traffic slightly drops when the \acrshort{grc} size is decreased.
If we return to Table~\ref{tab:spdc:results_overall}, in the previous chapter, in the columns for \acrshort{spdc}, we now understand why the total memory traffic is higher when the \acrshort{grc} size is reduced from 32 to 4 \glspl{set}.

Separating the inputs in a dedicated cache region also has the advantage of allowing prefetching techniques~\cite{Smith1982_Prefetcher} to be used efficiently, as prefetched cache-lines will not be undesirably evicted before use.
For the \acrshort{op} \acrshort{spmv} kernel, prefetching does not reduce the memory traffic, but it does improve performance by masking the main memory latency.

\begin{highlight}{Conclusion}{Red}
    To avoid unnecessary \glspl{eviction}, it is advantageous to \textbf{separate} the sequential reads of the inputs $A$ and $b$ from the ``random'' updates of the output $c$, into \textbf{dedicated cache regions}.
    A generic cache of reasonable size can \textbf{mask the latency} of the main memory when reading $A$ and $b$, and ensure this data is \textbf{read exactly once}.
\end{highlight}