% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% SpDCache %%
% Conclusion
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
\label{sec:spdc:conclu}

\lettrine{W}{e} have presented the concept of \acrlong{spdc}, a \gls{redC} that has optimized storage and compute techniques for the sparse and dense regions of \gls{banded} matrices.
This approach does not reduce the number of compute operations, however, it significantly decreases memory traffic (4.1\tm).
These benefits can be attributed to three techniques: (1) blocking the dense part of the \gls{band} in the \acrshort{drc} to take benefit of the existing spatial locality, (2) using fine-grained storage in the \acrshort{srt} to adapt to isolated elements, and (3) off-loading the \glspl{reduction} for isolated elements to the downstream caches to prevent local \glspl{eviction} and reduce traffic.

Exploiting the coarse-grain structure of matrices is key to designing hardware optimized for sparse matrix kernels.
In the next chapter, we present a more theoretical analysis of the behavior of \acrlong{spdc} before moving onto an actual \acrshort{rtl} implementation in Chapter~\ref{cha:arch}.

\bigskip\bigskip

\begin{highlight}{Takeaways}{Mulberry}
    \raggedright
    \colorbf{Mulberry}{Reduction cache:}
    \begin{itemize}
        \item Provides hardware support for performing \textbf{\glspl{reduction}}: $\mathit{RED}(key, value)$.
        \item The accumulations are done \textbf{near memory}, in the cache.
        \item \textbf{Reduces memory traffic} for kernels with a large number of \gls{reduction} updates.
    \end{itemize}

    \bigskip
    \colorbf{Mulberry}{SpDCache:}
    \begin{itemize}
        \item Considers that the matrix has a \textbf{dense region} and a less dense, \textbf{sparse region}. For \gls{banded} matrices, these correspond to the in-\gls{band} (\acrshort{ibr}) and the out-of-\gls{band} (\acrshort{obr}).
        \item Divides the L1 data-cache into \textbf{three regions}:
        \begin{itemize}
            \item \textbf{\acrshort{grc}}, generic region cache for regular memory accesses (such as reading $A$ and $b$);
            \item \textbf{\acrshort{drc}}, dense region cache for \glspl{reduction} in the dense region of the matrix;
            \item \textbf{\acrshort{srt}}, sparse region table for \glspl{reduction} in the sparse region of the matrix.
        \end{itemize}
        \item Uses memory transactions adapted to the region density: \textbf{cache-line-wise versus element-wise} for the \acrshort{drc} and the \acrshort{srt}, respectively.
        \item For a set of 48 matrices, results in \textbf{4.1\tms less memory traffic} for \acrshort{op} \acrshort{spmv} kernel.
    \end{itemize}
\end{highlight}

\clearpage