% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% SpDCache %%
% Generic Data-Cache
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Generic Data-Caches}
\label{sec:spdc:gc}

\lettrine{M}{odern} processors and accelerators integrate data-caches to bridge the latency and bandwidth gap between computation units and main memory~\cite{Hennessy2011_GC}.
A cache is a small, fast memory that stores recently accessed data anticipating that the data will be reused in the near future.
Its efficiency relies on two main forms of locality:
\begin{itemize}
\item temporal locality, where recently accessed data is likely to be reused within a short interval of time, and
\item spatial locality, where data stored near recently accessed addresses is likely to be used next.
\end{itemize}

\noindent
By exploiting these properties, caches reduce the average access latency and increase the overall bandwidth efficiency.

A cache is composed of several cache-lines, each grouping contiguous memory words.
The cache-line is the minimum granularity of data movement between memory levels.
When an access occurs, the cache controller searches for the requested address among the currently stored cache-lines.
If the address is found, a \gls{hit} occurs and the data is served directly to the core.
Otherwise, a \gls{miss} occurs, triggering a memory fetch from a lower cache level or from main memory.
If necessary, an existing cache-line is evicted to make room for the new data.
Each cache-line is associated with a \gls{tag} that records its address and additional bits to keep track of its state.

Several mapping policies exist.
In a direct-mapped cache, each memory block maps to a single cache-line.
This approach is simple to implement but results in conflicts and poor utilization.
A $n$-\gls{way} \gls{set}-associative cache allows each block to be placed in one of $n$ possible cache-lines within a \gls{set}, improving \gls{hit} rate with only a moderate additional hardware cost.
A fully-associative cache removes these constraints, as any block may occupy any cache-line, but requires more complex \gls{tag} searches.
In all cases, when a line must be evicted and multiple candidates exist, a replacement policy such as \acrfull{lru} or \emph{Random} decides which entry is evicted.

Caches have different policies when writes occur.
With a \emph{write-through} policy, every store operation updates both the cache and main memory, maintaining consistency but increasing traffic.
A \emph{write-back} policy delays the update until \gls{eviction}, reducing memory bandwidth at the cost of additional state bits (dirty flags) to maintain consistency.

For workloads dominated by \gls{seqAccesses}, these mechanisms and policies are highly effective.
However, \glspl{irrAccess} tend to degrade cache efficiency by increasing \gls{miss} rates and provoking frequent \glspl{eviction}.

Modern processors implement hierarchical cache organizations to balance latency, capacity, and throughput.
A typical CPU integrates small L1 caches close to each core, larger L2 caches shared by a group of cores, and a high-capacity L3 cache shared at the chip level.
Accelerators and domain-specific architectures often employ local data storage or scratchpad memories serving the same purpose, providing low-latency access to frequently reused data (see Chapter~\ref{cha:soa}).

The generic cache assumes regular read and write operations with predictable memory access patterns.
Yet workloads operating on sparse data violate this assumption.
In particular, ``random'' \glspl{reduction} often have poor performance because each \gls{reduction} triggers a read and a write of a full cache-line, to simply update one word.
Furthermore, the lack of spatial locality means each \gls{reduction} can trigger an \gls{eviction}, in the worst case.

In the following sections, we introduce cache architectures with native \gls{reduction} support, specifically designed to handle \gls{irrData}.
These architectures extend the classical cache model with \gls{reduction}-aware mechanisms, improving bandwidth efficiency under irregular access patterns.

\begin{highlight}{Conclusion}{Red}
    Generic data-caches are effective when access patterns exhibit spatial and temporal locality.
    This is not the case for sparse data and their performance is particularly poor for ``random'' \gls{reduction} operations.
    This limitation motivates the introduction of \textbf{\gls{reduction}-aware cache architectures} which are optimized for such ``random'' \glspl{reduction}.
\end{highlight}