% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% Microarchitecture %%
% Conclusion
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
\label{sec:arch:conclu}

\lettrine{I}{n} this chapter, we presented the microarchitectural implementation of \acrlong{spdc} and evaluated its performance through \acrshort{rtl} simulations.
By translating the theoretical insights from previous chapters into an actual hardware design, we demonstrated the effectiveness of \acrlong{spdc} in addressing the challenges of irregular memory access patterns in sparse matrix computations.

Our \acrshort{rtl} implementation of \acrlong{spdc}, with its three dedicated regions, \acrshort{grc}, \acrshort{drc}, and \acrshort{srt}, successfully optimizes memory transactions for both dense and sparse regions of matrices.
The results from \acrshort{rtl} simulations highlight significant reductions in memory traffic and improved cache utilization, validating the architectural choices and theoretical models developed earlier.
Specifically, \acrlong{spdc} achieves a substantial decrease in memory traffic compared to generic and \glspl{redC}, showcasing its ability to leverage the coarse-grain structure of sparse matrices.

Our findings also show the importance of fine-grain matrix structure.
The \acrshort{rtl} simulations revealed two distinct groups of matrices, split based on the $d_\mathit{cd}$ metric, which measures the density of non-zero elements within cache-line-sized portions of the matrices.
This distinction highlights how the fine-grain distribution of non-zero elements within matrices impacts performance.
By effectively handling both coarse-grain and fine-grain structure, \acrlong{spdc} demonstrates its adaptability across a diverse set of sparse matrices.

As a contribution, this chapter detailed the \acrshort{rtl} implementation of \acrlong{spdc}, including the hardware components and \gls{pipeline} stages that enable efficient data processing and \gls{reduction} operations.
Through \acrshort{rtl} simulations, we assessed \acrlong{spdc}'s performance, demonstrating its superiority in reducing memory traffic and improving cache utilization.
The \acrshort{rtl} results validated the theoretical models and insights, confirming the benefits of region-based caching and \gls{reduction} support.
Additionally, we identified and analyzed the impact of fine-grain matrix structure on performance, revealing two distinct groups of matrices based on the $d_\mathit{cd}$ metric.

However, several limitations must be acknowledged.
The current \acrshort{rtl} implementation is a first iteration and could be further optimized, as some features are still missing or not fully refined.
The use of element-wise, atomic transactions for offloading sparse data carries the risk of overwhelming the rest of the memory hierarchy if it is not designed to handle such transactions efficiently.
Finally, the evaluation was conducted on a single-core, single-cache level system, which helps us easily understand the cache behavior but limits raw performance compared to more complex, multi-level, or multi-core architectures, which are standard in high-performance computing systems.
However, we believe that with a more powerful processor (such as multiple-issue), the pressure on the memory system would be higher and the threshold between the two groups of matrices would be shifted to the right, with more matrices benefiting from \acrlong{spdc}.

Overall, this chapter highlights the critical role of both coarse-grain and fine-grain matrix structure in optimizing sparse matrix computations.
The contributions of \acrlong{spdc} not only advance the state-of-the-art in cache architectures but also provide a robust foundation for future enhancements in scalability and bandwidth efficiency, as discussed in the following chapter.
The ability to adapt to varying matrix structure ensures that \acrlong{spdc} remains effective across a wide range of applications and workloads.

\bigskip\bigskip

\begin{highlight}{Takeaways}{Mulberry}
    \colorbf{Mulberry}{Contributions:}
    \begin{itemize}
        \item \textbf{Microarchitectural Design:} We detailed the \acrshort{rtl} implementation of \acrlong{spdc}, including the hardware components and \gls{pipeline} stages that enable efficient data processing and \gls{reduction} operations.
        \item \textbf{Performance Evaluation:} Through \acrshort{rtl} simulations, we assessed \acrlong{spdc}'s performance, demonstrating its ability to reduce memory traffic and improve cache utilization.
        \item \textbf{Validation of Theoretical Models:} The \acrshort{rtl} results validated the theoretical models and insights, confirming the benefits of region-based caching and \gls{reduction} support.
        \item \textbf{Fine-Grain Structure Analysis:} We identified and analyzed the impact of fine-grain matrix structure on performance.
    \end{itemize}

    \bigskip
    \colorbf{Mulberry}{Limitations:}
    \begin{itemize}
        \item \textbf{\acrshort{rtl} Implementation Optimization:} The current \acrshort{rtl} implementation is a first iteration and could be further optimized, as some performance features are missing.
        \item \textbf{Offloading Risk:} The use of element-wise, atomic transactions for offloading sparse data carries the risk of overwhelming the downstream memory hierarchy.
        \item \textbf{Single-Core, Single-Cache Level System:} The evaluation was conducted on a single-core, single-cache level system, which is ideal for detailed analysis but limits performance compared to more recent architectures with multi-level caches and multi-core processors with multi-issue cores.
    \end{itemize}
\end{highlight}

\clearpage