% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% Microarchitecture %%
% Introduction
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{sec:arch:intro}

\lettrine{I}{n} the previous chapters, we established the theoretical foundations and \gls{highLevel} design principles for \acrlong{spdc}, a specialized cache architecture tailored to optimize memory transactions for sparse matrix computations.
This chapter bridges theory and practice by presenting the microarchitectural implementation of \acrlong{spdc} and evaluating its performance through \acrfull{rtl} simulations.

We begin by detailing the microarchitecture of \acrlong{spdc}, focusing on its three distinct regions: the \acrfull{grc}, the \acrfull{drc}, and the \acrfull{srt}.
Each region is designed to handle specific types of memory accesses, sequential reads, dense \glspl{reduction}, and sparse \glspl{reduction}, respectively, thereby addressing the irregular memory access patterns inherent in sparse matrix kernels.
The \acrshort{rtl} implementation of \acrlong{spdc} is described, highlighting the hardware components and \gls{pipeline} stages that enable efficient data processing and \gls{reduction} operations.

Following the microarchitectural overview, we present the evaluation methodology and the results obtained from \acrshort{rtl} simulations.
These simulations provide a detailed assessment of \acrlong{spdc}'s performance, including improvements in memory traffic and latency.
We analyze the behavior of \acrlong{spdc} processing a diverse set of sparse matrices, comparing its performance against baseline cache architectures and validating the theoretical insights gained from earlier chapters.

By translating the theoretical model into a practical hardware design, this chapter demonstrates the effectiveness of \acrlong{spdc} in minimizing memory traffic and improving computational efficiency for sparse matrix kernels.
The results underscore the importance of region-based caching and \gls{reduction} support, paving the way for further optimizations and scalability enhancements discussed in the following chapter.