% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% Optims %%
% Conclusion
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
\label{sec:optim:conclu}

\lettrine{T}{his} chapter explored optimizations and extensions to \acrlong{spdc}, focusing on enhancing its performance and scalability for sparse matrix computations.
We refined the \acrshort{rtl}-level implementation to maximize efficiency and resource utilization, ensuring \acrlong{spdc} remains robust and adaptable to modern computational demands.

We extended \acrlong{spdc} to matrix-matrix multiplication kernels, demonstrating its versatility beyond \acrshort{spmv}.
This extension highlights its adaptability to complex operations, reinforcing its potential as a foundational component for diverse sparse matrix computations.

A key focus was integrating \acrlong{spdc} into multi-level, multi-core architectures, addressing the need for scalable, distributed processing.
By optimizing memory traffic and parallel processing capabilities, \acrlong{spdc} becomes a viable solution for large-scale, high-performance applications.

While this chapter presents theoretical insights and proposed optimizations, no evaluation has been conducted yet.
The work aims to guide future research and implementation, with proper evaluation and analysis to conduct in subsequent studies.
These advancements position \acrlong{spdc} as a versatile and scalable solution for modern sparse matrix processing systems, paving the way for future research in high-performance architectures.

\bigskip\bigskip

\begin{highlight}{Takeaways}{Mulberry}
    \colorbf{Mulberry}{SpDCache Optimizations:}
    \begin{itemize}
        \item \textbf{Floating-Point Support:} Handle floating-point arithmetic within \acrlong{spdc};
        \item \textbf{Region-Size Configuration:} Methods to dynamically adjust region sizes in \acrlong{spdc};
        \item \textbf{\Gls{band}-Width Calculs in Hardware:} Hardware-based calculations for \acrshort{drc} \gls{band}-width;
        \item \textbf{Complete Inter-Region Cache Coherency:} Comprehensive cache coherency protocol for \acrlong{spdc}, ensuring data integrity and consistency across regions.
    \end{itemize}
    \bigskip
    \colorbf{Mulberry}{SpDCache on Matrix-Matrix Kernels:}
    \begin{itemize}
        \item \acrlong{spdc} effectively \textbf{extends to matrix-matrix kernels} through decomposition into matrix-vector operations;
        \item This demonstrates \acrlong{spdc}'s versatility across different sparse linear algebra kernels.
    \end{itemize}
    \bigskip
    \colorbf{Mulberry}{Multi-Level, Multi-Core SpDCache Architecture:}
    \begin{itemize}
        \item \acrlong{spdc} can be \textbf{scaled to multi-level, multi-core systems} by adapting slightly its architecture to each cache level and core;
        \item Vertical matrix partitioning enables efficient parallel processing across cores;
        \item The proposed architecture effectively addresses the challenges of scaling across cache levels and cores while optimizing memory usage.
    \end{itemize}
\end{highlight}

\clearpage