% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% Optims %%
% Introduction
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{sec:optim:intro}

\lettrine{T}{he} increasing complexity and scale of sparse matrix computations demand not only efficient single-core architectures but also scalable solutions that can leverage multi-core and multi-level memory hierarchies.
While the previous chapters have focused on the design, theoretical analysis, and microarchitectural implementation of \acrlong{spdc}, this chapter explores opportunities to further enhance its performance and scalability.

In Chapter~\ref{cha:arch}, we demonstrated the effectiveness of \acrlong{spdc} in reducing memory traffic and improving cache utilization for \acrlong{spmv} (\acrshort{spmv}).
However, as applications grow in size and complexity, the need for scalable architectures that can efficiently distribute workloads across multiple cores and memory levels becomes critical.
This chapter addresses these challenges by proposing optimizations and extensions to \acrlong{spdc} that target both single-core and multi-core systems.

We begin by discussing \acrshort{rtl}-level optimizations for \acrlong{spdc}, focusing on fine-tuning the hardware implementation to maximize performance and resource utilization.
Next, we explore the application of \acrlong{spdc} to \acrlong{spmspm} (\acrshort{spmspm}), extending its benefits beyond \acrshort{spmv} and demonstrating it can be adapted to a more complex kernel.

Finally, we investigate the integration of \acrlong{spdc} into multi-level, multi-core architectures.
This includes designing a scalable version of \acrlong{spdc} that can operate efficiently across multiple cores, as well as optimizing memory bandwidth usage to ensure high throughput in distributed environments.
By addressing these aspects, we aim to provide a comprehensive roadmap for scaling \acrlong{spdc} for high-performance sparse computation.

Through these optimizations and extensions, this chapter shows that the concepts of \acrlong{spdc} can be generalized and scaled to enhance performance of sparse matrix processing in real-world high-performance computing systems.

\clearpage