% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\FancyHeadRight{\sectitle}

\renewcommand{\sectitle}{Introduction}

\phantomsection
\chapter*{\sectitle}
\addstarredchapter{\sectitle}

% \setcounter{mtc}{10} %For the first minitoc. Position of first chapter that needs a minitoc.
% \minitoc

% \newcounter{introCnt}
% \setcounter{introCnt}{0}
% \renewcommand{\thesection}{\arabic{introCnt}}

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Intro 1

% \addtocounter{introCnt}{1}
% \section{Remove intro section title and minitoc if not needed}

%%%%%%%%%%%%%%%%%%%%

\note{From SoA paper}

The execution time of modern scientific and engineering applications is dominated by the manipulation of large sparse matrices, \emph{i.e.} matrices whose number of non-zero entries $NNZ$ is much smaller than their dimensions $M \times N$.
Matrices derived from computational physics problems are often sparse and regular because they derive from a relatively small number of differential terms.
Algebraic graph problems produce matrices which are still sparse but usually less regular~\cite{Buluc2011}.
Similarly, the weight and activation matrices used by Convolutional Neural Networks (CNN) may hold millions of terms, but most of them are explicitly set to zero (\emph{e.g.} \emph{rectified}) without loss of information~\cite{Gale2020}.
It was recognized early in the 1960s~\cite{Saad2003_COO_CSR_DIA_ELL} that working on these sparse data sets using dense array representations was inefficient.
The seminal work of OSKI~\cite{Vuduc2005} in 2003 formalized the sparse representation alternatives and identified many optimizations.
This opened the path for specialized software libraries for sparse algebra~\cite{Williams2007}.
However, the single-core performance of software implementations can be deceiving, often only reaching 10\% of peak performance~\cite{Gale2020}.

%%%% VC %%%%
The main matrix multiplication kernels that are of interest are matrix-vector (MV) and matrix-matrix (MM), which include \acrshort{spmv} and SpMSpV ($A \times b = c$), SpGEMV ($A \times b + d = c$), SpMM and SpMSpM ($A \times B = C$), and SpGEMM ($A \times B + D = C$), with \emph{Sp} for \emph{sparse} and \emph{GE} for \emph{general}.
%%%% VL %%%%
% The main operations that are of interest are SpMV, SpMSpV, GEMV, SpGEMV and their equivalent with matrix-matrix inputs as shown in Table~\ref{tab:kernels}.
%
% \begin{table}[htbp]
%     \caption{Main Matrix Multiplication Kernels}
%     \begin{center}
%         \begin{tabular}{|c||*{3}{c|}|*{2}{c|}}
%             \hhline{-||---||--}
%             \textit{\textbf{Name}} & \textbf{MV} & \textbf{SpMV} & \textbf{SpMSpV} & \textbf{GEMV} & \textbf{SpGEMV} \\
%             \hhline{-||---||--}
%             \textit{Formula} & \multicolumn{3}{c||}{$A \times b = c$} & \multicolumn{2}{c|}{$A \times b + d = c$} \\
%             \hhline{-||---||--}
%             \emph{A} & dense & sparse & sparse & dense & sparse \\
%             \hhline{-||---||--}
%             \emph{b} & dense & dense & sparse & dense & dense \\
%             \hhline{=::===::==}
%             \textit{\textbf{Name}} & \textbf{MM} & \textbf{SpMM} & \textbf{SpMSpM} & \textbf{GEMM} & \textbf{SpGEMM} \\
%             \hhline{-||---||--}
%             \textit{Formula} & \multicolumn{3}{c||}{$A \times B = C$} & \multicolumn{2}{c|}{$A \times B + D = C$} \\
%             \hhline{-||---||--}
%             \emph{A} & dense & sparse & sparse & dense & sparse \\
%             \hhline{-||---||--}
%             \emph{B} & dense & dense & sparse & dense & sparse \\
%             \hhline{-||---||--}
%         \end{tabular}
%         \label{tab:kernels}
%     \end{center}
% \end{table}
%%%%%%%%%%%%
SpMV, as well as its variant SpGEMV, are by far the dominant operations of modern algebraic kernels (linear solvers, eigensolvers) which have been explicitly based on them~\cite{Gilbert2008, Duff2002, Penn2006, Yamazaki2011, Itoh1995, Kjolstad2017_TACO, Hapla2013}. Since these kernels are ubiquitous, their efficient implementation is primordial.
In contrast, the SpMSpM multiplication is less frequent in standard algebraic kernels, but it is extensively used in algebraic graph manipulation~\cite{Dongen2000, Kepner2015, Gilbert2007, Rabin1989, Chan2007, Azad2015, Brin1998}, as well as in multigrid solvers~\cite{Brandt2003, Li2021, Ryabkov2022, Ayyar2023}. The growing importance of large graphs has motivated new research on optimizing SpMSpM applications. 

The two matrix operations, SpMV and SpMSpM, pose orthogonal computational challenges: (1) access to sparse input data, (2) \emph{reduction} operations on the intermediate result and (3) formatting and writing back the result to memory. The different algorithms, which are detailed in Section~\ref{sec:bg:matrix_formats_and_algos}, are built on different articulations of these three aspects.
The sparse data storage scheme used in memory dictates the first aspect and is briefly explored in Section~\ref{sec:bg:matrix_formats}.
%The sparse data storage scheme used in memory dictates the first aspect and we explore different alternatives in Subsection~\ref{subsec:matrix_formats}.
The third aspect, \emph{i.e.} result storage, is sometimes overlooked when the output is dense and not too large. For example, when running on a conventional CPU which is memory-bound, it becomes crucial to streamline the write accesses in order to optimize cache usage when writing a large dense vector output.
The second aspect, \emph{i.e.} the \emph{reduction} operations, deserves different solutions according to the execution platform. On a conventional CPU, the bottleneck is memory throughput. Thus, most software implementations of SpMV deal with \emph{a priori} reorganization of the input matrices, by row ordering, blocking and loop optimization, for improving the data locality of operations.
The perspective for SpMV is different when dealing with GPU platforms. The challenge shifts to finding a balanced assignment of the numerous threads and warps, and to the use of local memory resources~\cite{Liu2014}.
SpMSpM is far more complex: the intermediate \emph{reductions} introduce new $NNZ$ values at random locations, which trigger costly memory allocations and list management operations~\cite{Gao2020}. As a consequence, this phase dominates the execution time and becomes the main memory bottleneck on a regular CPU. The issue is similar on GPUs because of the limited size of the scratchpad memory for processing large matrices, which requires complex memory management and load balancing~\cite{Zachariadis2020, Nagasaka2018, Deveci2018}.

The main motivation to use custom hardware accelerators is to make better use of the intermediate memory hierarchy, ensuring the input data is available near the cores and also off-loading the accumulation and \emph{reduction} operations from the processors. The focus of this paper is to review these architectures, to compare them and to analyze their performance. Our main focus is on accelerators that eventually target custom or ASIC implementations, although we do touch on the contributions from accelerators that were prototyped on FPGAs. The reader is referred to~\cite{MOHAMMED_2022_SPARSE} for a survey of acceleration techniques for CPUs, GPUs and FPGAs. Section~\ref{sec:bg:matrix_formats_and_algos} reviews the storage schemes and the three basic algorithms with a focus on the data access patterns and their impact on the memory hierarchy. In Section~\ref{sec:bg:hw_challenges} we study how the different algorithms stress the memory systems and propose an analytical model based on these insights. In Section~\ref{sec:soa:hw_accelerators} we present the operation of several state-of-the-art hardware accelerators, in the light of these hardware challenges. In Section~\ref{sec:soa:comparison_and_analysis}, we systematically compare these accelerators. Throughout this study, we try to be consistent, and we hope that our nomenclature will be useful for architects and designers who intend to implement dedicated hardware for problems with sparse datasets.

%%%%%%%%%%%%%%%%%%%%


\bigskip\bigskip\bigskip\bigskip\bigskip\bigskip

\lettrine{T}{he} rapid growth of data-intensive applications has made sparse and irregular data structures increasingly prevalent in scientific computing, machine learning, and graph analytics.
However, efficiently processing sparse matrices remains a fundamental challenge due to their inherent irregularity, which leads to unpredictable memory access patterns, limited opportunities for data reuse, and suboptimal utilization of hardware resources.
Existing hardware accelerators have demonstrated partial solutions, yet they often rely on specialized assumptions, exhibit limited scalability, or fail to fully exploit reduction and memory access locality.

This thesis addresses the following central question:

\bigskip
\begin{highlight}{Problematic}{Mulberry}
    \colorbf{Mulberry}{\emph{How can hardware and architectural mechanisms be designed to efficiently support the computation of irregular sparse data, in particular random reductions, while ensuring scalability, adaptability, and performance across diverse workloads?}}
\end{highlight}
\bigskip

To tackle this problem, we investigate the limitations of existing algorithms and accelerators, propose a novel cache architecture with reduction support tailored to sparse data, and evaluate its behavior through modeling, simulation, and hardware prototyping.

\bigskip\bigskip

This manuscript is structured around several key contributions.
We begin with an in-depth discussion of irregular data, with a particular emphasis on sparse matrices (Chapter~\ref{cha:background}).
This includes a study of fundamental sparse matrix multiplication kernels and their associated algorithms, highlighting the main hardware challenges that arise when dealing with sparse data.
A high-level analysis of algorithms for sparse data computing is provided, with a focus on basic storage schemes and opportunities for data reuse.
Building on this foundation, we review the state-of-the-art in hardware accelerators designed for efficient sparse matrix computing (Chapter~\ref{cha:soa}).
This review identifies common characteristics, proposes a classification of existing accelerators and design tendencies, and offers a comparative analysis to better understand the rationale behind specific architectural choices under different objectives and constraints.

Following this survey, we introduce a novel cache architecture specifically designed to address the challenge of “random” reductions (Chapter~\ref{cha:spdc}).
The proposed cache integrates reduction support and organizes its memory into multiple regions tailored to denser and sparser data patterns.
A Python model was developed to simulate memory transactions in this cache and to compare its behavior against that of a generic cache.
To further investigate its performance, we complement this analysis with a probabilistic model, implemented in C, that enables rapid simulation and design space exploration (Chapter~\ref{cha:theory}).
At the micro-architectural level, we detail the virtual partitioning of cache memory into multiple configurable regions, which ensures both flexibility and adaptability to diverse workloads (Chapter~\ref{cha:microarch}).

The proposed cache is then evaluated through RTL simulation, which reveals two distinct groups of matrices that exhibit different performance profiles (Chapter~\ref{cha:eval}).
Finally, the manuscript outlines potential future enhancements to improve both performance and scalability, including a multicore extension of the cache architecture and optimizations targeting bandwidth efficiency (Chapter~\ref{cha:optim}).

\section*{Contributions of the Thesis}

The main contributions of this thesis can be summarized as follows:

\begin{enumerate}
    \item \textbf{Analysis of irregular and sparse data computing} {\raggedleft \hfill (Chapter~\ref{cha:background})}
    \begin{itemize}
        \item Study of fundamental sparse matrix multiplication kernels and algorithms.
        \item Identification of the key hardware challenges associated with sparse data processing.
        \item High-level analysis of algorithms for sparse data computing, with emphasis on storage schemes and data reuse opportunities.
    \end{itemize}

    \item \textbf{Survey and classification of state-of-the-art accelerators} {\raggedleft \hfill (Chapter~\ref{cha:soa})}
    \begin{itemize}
        \item Comprehensive review of hardware accelerators for sparse matrix computing.
        \item Identification of common architectural features and design tendencies.
        \item Comparative analysis to explain design choices under varying objectives and constraints.
    \end{itemize}

    \item \textbf{Design of a cache architecture for random reductions} {\raggedleft \hfill (Chapter~\ref{cha:spdc})}
    \begin{itemize}
        \item Proposal of a cache with native reduction support, capable of handling both dense and sparse data via multiple cache regions.
        \item Development of a Python-based simulation model to analyze memory transactions and benchmark against a generic cache.
    \end{itemize}

    \item \textbf{Probabilistic modeling and exploration of cache behavior} {\raggedleft \hfill (Chapter~\ref{cha:theory})}
    \begin{itemize}
        \item Implementation of a lightweight C-based probabilistic model for rapid simulation.
        \item Exploration of cache dimensioning and behavior across diverse workloads.
    \end{itemize}

    \item \textbf{Micro-architecture of the reduction cache} {\raggedleft \hfill (Chapter~\ref{cha:microarch})}
    \begin{itemize}
        \item Introduction of a virtual memory partitioning scheme enabling multiple configurable regions.
        \item Emphasis on flexibility and adaptability in cache design.
    \end{itemize}

    \item \textbf{Evaluation through RTL simulation} {\raggedleft \hfill (Chapter~\ref{cha:eval})}
    \begin{itemize}
        \item Assessment of the proposed cache architecture using RTL-level simulation.
        \item Identification of two distinct groups of sparse matrices leading to different performance profiles.
    \end{itemize}

    \item \textbf{Future directions for performance and scalability} {\raggedleft \hfill (Chapter~\ref{cha:optim})}
    \begin{itemize}
        \item Proposal of a multicore version of the cache to extend scalability.
        \item Exploration of bandwidth optimization strategies.
    \end{itemize}
\end{enumerate}

% \subsection{A Section}

% \subsubsection{A Subsection}

% \lipsum[3]

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Intro 2

% \addtocounter{introCnt}{1}
% \section{Second element}

% \lipsum[1-5]

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \renewcommand{\thesection}{\arabic{chapter}.\arabic{section}}

\cleardoublepage %% Comment this line if this section occupies the last pages