% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Resume and Abstract
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if\Language\EN

\FancyHeadRight{\sectitle}

\renewcommand{\sectitle}{Abstract}

\phantomsection
\addcontentsline{toc}{chapter}{\sectitle}
\chapter*{\sectitle}

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Your Text %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\lettrine{S}{parse} matrices are ubiquitous in data-intensive applications such as scientific computing, graph analytics, and machine learning, yet their irregular memory access patterns pose significant challenges for traditional computing architectures.
This thesis addresses the core challenge of designing hardware and architectural mechanisms to efficiently support sparse data processing, focusing on scalability, adaptability, and high performance.

We analyze the fundamental algorithms for sparse matrix multiplication, \acrlong{ip}, \acrlong{op}, and \acrlong{gust}'s, and identify their hardware challenges, separated into the \gls{gather} and \gls{scatter} problems.
\acrlong{gust}'s algorithm emerges as the most promising for hardware implementation due to its balanced trade-offs, while the \acrlong{op} algorithm is well-suited for \acrlong{mv} kernels.

Our main contribution is \acrlong{spdc}, a region-based \acrlong{redC} architecture that optimizes memory transactions for sparse data by leveraging the structural properties of \gls{banded} matrices.
\acrlong{spdc} partitions cache memory into sparse and dense regions, significantly reducing memory traffic compared to generic caches.
Key contributions include:
\begin{itemize}
    \item Architectural Design and Modeling: A Python-based simulation model and a probabilistic C-based model analyze \acrlong{spdc}'s memory transactions, revealing the importance of coarse-grain matrix structure (\ex \gls{banded} matrices).
    \item Microarchitectural Implementation: \acrshort{rtl}-level simulations validate \acrlong{spdc}'s effectiveness in reducing memory traffic and improving cache utilization, with virtual partitioning ensuring flexibility and adaptability.
    Further analysis reveals the importance of fine-grain matrix structure in optimizing performance (\eg cache-line density).
    \item Optimizations and Scalability: Extensions to sparse \acrlong{mm} kernels and multi-core architectures position \acrlong{spdc} as a versatile, scalable solution for modern sparse matrix processing.
\end{itemize}

This thesis advances the state-of-the-art in cache architectures for sparse data, emphasizing intrinsic understanding and adaptive, structure-aware design for high-performance hardware solutions.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\cleardoublepage %% Comment this line if this section occupies the last pages
\adjustmtc

\fi

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{otherlanguage}{french}

\FancyHeadRight{\sectitle}

\renewcommand{\sectitle}{R\'esum\'e en Fran\c cais}

\phantomsection
\addcontentsline{toc}{chapter}{\sectitle}
\chapter*{\sectitle}

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Your Text %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\lettrine{L}{es} matrices creuses sont omniprésentes dans les applications gourmandes en données, telles que le calcul scientifique, l'analyse de graphes et l'apprentissage automatique.
Pourtant, leurs schémas d'accès mémoire irréguliers posent des défis majeurs pour les architectures d'ordinateurs et de mémoires traditionnelles.
Cette thèse aborde le défi central de la conception de mécanismes matériels et architecturaux pour prendre en charge efficacement le traitement des données creuses, en mettant l'accent sur la scalabilité, l'adaptabilité et les hautes performances.

Nous analysons les algorithmes fondamentaux de multiplication de matrices creuses, à savoir \acrlong{ip}, \acrlong{op} et \acrlong{gust}, et identifions leurs défis matériels, séparés en problèmes de \gls{gather} (recherche d'intersections) et de \gls{scatter} (réductions aléatoires).
L'algorithme de \acrlong{gust} s'impose comme le plus prometteur pour une implémentation matérielle en raison de ses compromis équilibrés, tandis que l'algorithme \acrlong{op} est particulièrement adapté aux opérations matrice-vecteur.

Notre contribution principale est \acrlong{spdc}, une architecture de cache de réduction basée sur des régions qui optimise les transactions mémoire pour les données creuses en exploitant les propriétés structurelles des matrices bandées.
\acrlong{spdc} partitionne la mémoire cache en régions denses et creuses, réduisant significativement le trafic mémoire par rapport aux caches génériques.
Les contributions clés incluent :
\begin{itemize}
    \item Conception architecturale et modélisation : Un modèle de simulation basé sur Python et un modèle probabiliste en C analysent les transactions mémoire de \acrlong{spdc}, révélant l'importance de la structure matricielle à gros grain (par exemple, les matrices bandées).
    \item Implémentation microarchitecturale : Des simulations au niveau \acrshort{rtl} valident l'efficacité de \acrlong{spdc} dans la réduction du trafic mémoire et l'amélioration de l'utilisation du cache, avec un partitionnement virtuel de la mémoire garantissant flexibilité et adaptabilité.
    Une analyse approfondie révèle l'importance de la structure matricielle à grain fin pour optimiser les performances (en particulier la densité des lignes de cache).
    \item Optimisations et scalabilité : Des extensions aux multiplications matrice-matrice et aux architectures multi-c\oe{}urs positionnent \acrlong{spdc} comme une solution polyvalente et évolutive pour le traitement moderne des matrices creuses.
\end{itemize}

Cette thèse fait progresser l'état-de-l'art des architectures de cache pour les données creuses, en mettant l'accent sur une compréhension intrinsèque et une conception adaptative et sensible à la structure pour des solutions matérielles hautes performances.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{otherlanguage}
\cleardoublepage %% Comment this line if this section occupies the last pages
\adjustmtc